F\# Data: \textnormal{Making structured data first-class citizens}

### 2015-02-27 icfp-2015/paper.tex
Use \nu? rather than \nu_opt for optional record names

The grammar below defines \emph{structural type} $\sigma$. We distinguish between \emph{non-nullable types}
that always have a valid value (written as $\hat{\sigma}$) and \emph{nullable types} that encompass missing 
and \kvd{null} values (written as $\sigma$). We write $\nu$ for record field names and $\nu_{\ident{opt}}$
for record names, which can be empty:
%
\begin{equation*}
\begin{array}{rcl}
 \hat{\sigma} &\narrow{=}& \nu_{\ident{opt}}\; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \\[0.1em]
                &\narrow{|}& \ident{float} \lsep \ident{decimal} \lsep \ident{int} \lsep \ident{bit} \lsep \ident{bool} \lsep \ident{string} 
 \\[0.6em] 
       \sigma &\narrow{=}& ~\hat{\sigma}~ \lsep \hat{\sigma}\;\,\kvd{option} \lsep [\sigma] \\[0.1em]
              &\narrow{|}& \sigma_1 + \ldots + \sigma_n \lsep ~\top~ \lsep \kvd{null}
\end{array}
\end{equation*}
%
The non-nullable types include records (consisting of an optional name and zero or more fields with
their types) and primitive types. Records arising from XML documents are always named, while records
used by JSON and CSV providers are always unnamed.

@@

\noindent\boxed{\emph{Record types}}
\begin{align}
\hspace{-1em}
\label{eq:sub-record1}\tag{R1}
 \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} :> 
 \nu~\{ \nu_1\!:\!\sigma_1', .., \nu_n\!:\!\sigma_n' \}
 \;\; (\sigma_i :> \sigma_i')
\end{align}
\vspace{-1.4em}
\begin{align}
\hspace{-1em}
\label{eq:sub-record2}\tag{R2}
 \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} :> 
 \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_m\!:\!\sigma_m \}
 \;\; (m \geq n)
\end{align}
\vspace{-1.0em}
\begin{align}
\label{eq:sub-record3}\tag{R3}
\hspace{-1.75em}
\begin{array}{l}
 \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} :> \\
 \nu~\{ \nu_{\pi(1)}\!:\!\sigma_{\pi(1)}, .., \nu_{\pi(m)}\!:\!\sigma_{\pi(m)} \}
\end{array} \qquad &(\textnormal{permutation}~\pi)
\end{align}
\vspace{-0.6em}
\begin{align}
\label{eq:sub-record4}\tag{R4}
\hspace{-1.75em}
\begin{array}{l}
 \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n, \nu_{n+1}\!:\!\kvd{null}, .., \nu_{n+m}\!:\!\kvd{null} \} :> \\
 \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \}
\end{array}
\end{align}
\end{definition}

@@

\begin{figure*}[t]
\noindent
\begin{equation*}
\hspace{3em}
\inference[(record-1)\;]
  { (\nu_i = \nu'_j \Leftrightarrow (i = j) \wedge (i \leq k))
      \qquad \forall i\in\{ 1 .. k \}.(\sigma_i \tsep \sigma'_i \vdash \sigma''_i) }
  { \begin{array}{l}
    \nu \; \{ \nu_1 : \sigma_1,  \; \ldots \;, \nu_k : \sigma_k, \; \ldots \;, \nu_n : \tau_n \} \tsep
    \nu \; \{ \nu'_1 : \sigma'_1, \; \ldots \;, \nu'_k : \sigma'_k, \; \ldots \;, \nu'_m : \tau'_m \} \vdash\\
    \nu \; \{ \nu_1 : \sigma''_1, \; \ldots \; , \nu_k : \sigma''_k, 
                            \nu_{k+1} : \addopt{\sigma_{k+1}}, \ldots, \nu_n : \addopt{\sigma_n},
                            \nu'_{k+1} : \addopt{\sigma'_{k+1}}, \ldots, \nu'_m : \addopt{\sigma'_m} \}
    \end{array} }
\end{equation*}

@@

\begin{equation*}
\hspace{2em}
\inference[(order)\;]
  { \nu \; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \tsep \sigma \vdash \sigma' }
  { \nu \; \{ \nu_{\pi(1)} : \sigma_{\pi(1)}, \ldots, \nu_{\pi(n)} : \sigma_{\pi(n)} \} \tsep \sigma \vdash \sigma' }
\quad
\inference
  { \sigma_1 + \ldots + \sigma_n \tsep \sigma \vdash \sigma' }
  { \sigma_{\pi(1)} + \ldots + \sigma_{\pi(n)} \tsep \sigma \vdash \sigma' }
\qquad (\pi~\textnormal{permutation})  
\end{equation*}

@@

As suggested by subtyping, our system permits certain runtime conversions (\emph{e.g.}~a numerical 
value \num{1} can be treated as \kvd{true}). These are captured by the following operations:
%
\begin{equation*}
\begin{array}{lcl}
 op &\narrow{=}& \ident{asInt}(s) \lsep \ident{asDec}(s) \lsep \ident{asFloat}(s) \lsep \ident{asBool}(s)  \\
    &\narrow{|}& \ident{asStr}(s) \lsep \ident{getChildren}(s) \lsep \ident{getField}(\nu_{\ident{opt}}, \nu, s)\\
    &\narrow{|}& \ident{isNull}(s) \lsep \ident{isTag}(\ident{tag}, s)
\end{array}
\end{equation*}
%
The reduction rules for these operations are defined in Figure~\ref{fig:op-conversions}. It is important
to note that the relativised property we prove about F\# Data specifies \emph{sufficient}, but
not \emph{necessary} conditions. That is, a type provided based on a sample is guaranteed to work 
correctly on certain inputs, but may also handle several additional inputs.

@@

\begin{figure*}
\begin{multicols}{2}
% Primitive structural types become corresponding F# types and conversion is inserted
\noindent
\begin{equation*}
\begin{array}{l}
 \sem{\sigma_p}_e = \tau_p,op(e),\emptyset\quad\textnormal{where} \\[0.4em]
\quad\sigma_p, \tau_p, op\in  \{~ (\ident{bit}, \ident{bool}, \ident{asBool}),  (\ident{bool}, \ident{bool}, \ident{asBool})\\[0.2em]
\hspace{2.9em} (\ident{int}, \ident{int}, \ident{asInt}), (\ident{decimal},\ident{decimal},\ident{asDec}),\\[0.2em]
\hspace{2.9em} (\ident{float},\ident{float},\ident{asFloat}), (\ident{string},\ident{string},\ident{asStr}) ~\}\\[0.6em]
\end{array}
\end{equation*}
%
% Records become classes
\begin{equation*}
\begin{array}{l}
 \sem{\nu_{\ident{opt}}\; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \}}_e = \\[0.1em]
 \quad C, \kvd{new}~C(e), L_1\cup\ldots\cup L_n\cup\{ L \}\quad\textnormal{where}\\[0.6em]
 \qquad \;\;C~\textnormal{is a fresh class name} \\[0.1em]
 \qquad \;\,\,L = \kvd{type}~C(v:\ident{StructVal})~=~M_1\ldots M_n  \\[0.1em]
 \qquad M_i = \kvd{member}~\nu_i:\tau_i=e_i\\[0.1em]
 \qquad \tau_i, e_i, L_i = \sem{\sigma_i}_{e'},\;\; e'=\ident{getField}(\nu_{\ident{opt}}, \nu_i, e)\\[0.6em]
\end{array}
\end{equation*}
%
% Collections
\begin{equation*}
\begin{array}{l}
 \sem{\,[\sigma]\,}_e = \ident{list}\langl\tau\rangl, \ident{List.map}~(\lambda x\rightarrow e')~(\ident{getChildren}(e)), L\\[0.4em]
 \quad \textnormal{where}~~\tau, e', L = \sem{\hat{\sigma}}_x
\end{array}
\end{equation*}

@@

Given a JSON, XML or CSV document, the F\# Data implementation constructs a structural value $s$
from the sample (this is straightforward, but Section~\ref{sec:impl-parsing} discusses some interesting 
aspects). The following defines a mapping $\semalt{-}$ which turns a sample value $s$ into a structural
type $\sigma$:
%
\begin{equation*}
\begin{array}{rclcrcl}
 \semalt{0} &\narrow{=}& \ident{bit} &\qquad& \semalt{1} &\narrow{=}& \ident{bit} \\
 \semalt{i} &\narrow{=}& \ident{int} && \semalt{d} &\narrow{=}& \ident{decimal} \\
 \semalt{f} &\narrow{=}& \ident{float} && \semalt{t} &\narrow{=}& \ident{string} \\
 \semalt{b} &\narrow{=}& \ident{bool} && \semalt{\kvd{null}}  &\narrow{=}& \kvd{null} \\[0em]
\end{array}
\end{equation*}
\noindent
\vspace{-0.2em}
\begin{equation*}
\begin{array}{l}
 \semalt{[s_1; \ldots; s_n]} = [\semalt{s_1, \ldots, s_n}]
 \\[0.5em]
 \semalt{\nu_{\ident{opt}}~\{ \nu_1 \mapsto s_1, \ldots, \nu_n \mapsto s_n \}} =\\[0.1em]
 \qquad\nu_{\ident{opt}}~\{ \nu_1:\semalt{s_1}, \ldots, \nu_n :\semalt{s_n} \}
 \\[0.5em]
 \semalt{s_1, \ldots, s_n} = \sigma_n \quad\textnormal{where}\\[0.1em]
 \qquad\sigma_0 = \top,~ \forall i\in \{ 1.. n \}.~ \sigma_{i-1} \triangledown \semalt{s_i} \vdash \sigma_i
\end{array}
\end{equation*}
%
Primitive values are mapped to their corresponding types, with the exception of $0$ and $1$, which
are inferred as \ident{bit}. For records, we return a type with field types inferred based on 
individual values.

>>>>

The grammar below defines \emph{structural type} $\sigma$. We distinguish between \emph{non-nullable types}
that always have a valid value (written as $\hat{\sigma}$) and \emph{nullable types} that encompass missing 
and \kvd{null} values (written as $\sigma$). We write $\nu$ for record field names and $\nu?$
for record names, which can be empty:
%
\begin{equation*}
\begin{array}{rcl}
 \hat{\sigma} &\narrow{=}& \nu?\; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \\[0.1em]
                &\narrow{|}& \ident{float} \lsep \ident{decimal} \lsep \ident{int} \lsep \ident{bit} \lsep \ident{bool} \lsep \ident{string} 
 \\[0.6em] 
       \sigma &\narrow{=}& ~\hat{\sigma}~ \lsep \hat{\sigma}\;\,\kvd{option} \lsep [\sigma] \\[0.1em]
              &\narrow{|}& \sigma_1 + \ldots + \sigma_n \lsep ~\top~ \lsep \kvd{null}
\end{array}
\end{equation*}
%
The non-nullable types include records (consisting of an optional name and zero or more fields with
their types) and primitive types. Records arising from XML documents are always named, while records
used by JSON and CSV providers are always unnamed.

@@

\noindent\boxed{\emph{Record types}}
\begin{align}
\hspace{-1em}
\label{eq:sub-record1}\tag{R1}
 \nu?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \hspace{-0.1em}:> \hspace{-0.1em}
 \nu?~\{ \nu_1\!:\!\sigma_1', .., \nu_n\!:\!\sigma_n' \}
 \; (\sigma_i \hspace{-0.25em}:>\hspace{-0.25em} \sigma_i')
\end{align}
\vspace{-1.4em}
\begin{align}
\hspace{-1em}
\label{eq:sub-record2}\tag{R2}
 \nu?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \hspace{-0.1em} :> \hspace{-0.1em}
 \nu?~\{ \nu_1\!:\!\sigma_1, .., \nu_m\!:\!\sigma_m \}
 \; (m \hspace{-0.1em}\geq\hspace{-0.1em} n)
\end{align}
\vspace{-1.0em}
\begin{align}
\label{eq:sub-record3}\tag{R3}
\hspace{-1.75em}
\begin{array}{l}
 \nu?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} :> \\
 \nu?~\{ \nu_{\pi(1)}\!:\!\sigma_{\pi(1)}, .., \nu_{\pi(m)}\!:\!\sigma_{\pi(m)} \}
\end{array} \qquad &(\textnormal{permutation}~\pi)
\end{align}
\vspace{-0.6em}
\begin{align}
\label{eq:sub-record4}\tag{R4}
\hspace{-1.75em}
\begin{array}{l}
 \nu?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n, \nu_{n+1}\!:\!\kvd{null}, .., \nu_{n+m}\!:\!\kvd{null} \} :> \\
 \nu?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \}
\end{array}
\end{align}
\end{definition}

@@

\begin{figure*}[t]
\noindent
\begin{equation*}
\hspace{3em}
\inference[(record-1)\;]
  { (\nu_i = \nu'_j \Leftrightarrow (i = j) \wedge (i \leq k))
      \qquad \forall i\in\{ 1 .. k \}.(\sigma_i \tsep \sigma'_i \vdash \sigma''_i) }
  { \begin{array}{l}
    \nu? \; \{ \nu_1 : \sigma_1,  \; \ldots \;, \nu_k : \sigma_k, \; \ldots \;, \nu_n : \tau_n \} \tsep
    \nu? \; \{ \nu'_1 : \sigma'_1, \; \ldots \;, \nu'_k : \sigma'_k, \; \ldots \;, \nu'_m : \tau'_m \} \vdash\\
    \nu? \; \{ \nu_1 : \sigma''_1, \; \ldots \; , \nu_k : \sigma''_k, 
                            \nu_{k+1} : \addopt{\sigma_{k+1}}, \ldots, \nu_n : \addopt{\sigma_n},
                            \nu'_{k+1} : \addopt{\sigma'_{k+1}}, \ldots, \nu'_m : \addopt{\sigma'_m} \}
    \end{array} }
\end{equation*}

@@

\begin{equation*}
\hspace{1em}
\inference[(order-1)\;]
  { \nu? \; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \tsep \sigma \vdash \sigma' }
  { \nu? \; \{ \nu_{\pi(1)} : \sigma_{\pi(1)}, \ldots, \nu_{\pi(n)} : \sigma_{\pi(n)} \} \tsep \sigma \vdash \sigma' }
\quad
\inference[(order-2)\;]
  { \sigma_1 + \ldots + \sigma_n \tsep \sigma \vdash \sigma' }
  { \sigma_{\pi(1)} + \ldots + \sigma_{\pi(n)} \tsep \sigma \vdash \sigma' }
\qquad (\pi~\textnormal{permutation})  
\end{equation*}

@@

As suggested by subtyping, our system permits certain runtime conversions (\emph{e.g.}~a numerical 
value \num{1} can be treated as \kvd{true}). These are captured by the following operations:
%
\begin{equation*}
\begin{array}{lcl}
 op &\narrow{=}& \ident{asInt}(s) \lsep \ident{asDec}(s) \lsep \ident{asFloat}(s) \lsep \ident{asBool}(s)  \\
    &\narrow{|}& \ident{asStr}(s) \lsep \ident{getChildren}(s) \lsep \ident{getField}(\nu?, \nu, s)\\
    &\narrow{|}& \ident{isNull}(s) \lsep \ident{isTag}(\ident{tag}, s)
\end{array}
\end{equation*}
%
The reduction rules for these operations are defined in Figure~\ref{fig:op-conversions}. It is important
to note that the relativised property we prove about F\# Data specifies \emph{sufficient}, but
not \emph{necessary} conditions. That is, a type provided based on a sample is guaranteed to work 
correctly on certain inputs, but may also handle several additional inputs.

@@

\begin{figure*}
\begin{multicols}{2}
% Primitive structural types become corresponding F# types and conversion is inserted
\noindent
\begin{equation*}
\begin{array}{l}
 \sem{\sigma_p}_e = \tau_p,op(e),\emptyset\quad\textnormal{where} \\[0.4em]
\quad\sigma_p, \tau_p, op\in  \{~ (\ident{bit}, \ident{bool}, \ident{asBool}),  (\ident{bool}, \ident{bool}, \ident{asBool})\\[0.2em]
\hspace{2.9em} (\ident{int}, \ident{int}, \ident{asInt}), (\ident{decimal},\ident{decimal},\ident{asDec}),\\[0.2em]
\hspace{2.9em} (\ident{float},\ident{float},\ident{asFloat}), (\ident{string},\ident{string},\ident{asStr}) ~\}\\[0.6em]
\end{array}
\end{equation*}
%
% Records become classes
\begin{equation*}
\begin{array}{l}
 \sem{\nu?\; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \}}_e = \\[0.1em]
 \quad C, \kvd{new}~C(e), L_1\cup\ldots\cup L_n\cup\{ L \}\quad\textnormal{where}\\[0.6em]
 \qquad \;\;C~\textnormal{is a fresh class name} \\[0.1em]
 \qquad \;\,\,L = \kvd{type}~C(v:\ident{StructVal})~=~M_1\ldots M_n  \\[0.1em]
 \qquad M_i = \kvd{member}~\nu_i:\tau_i=e_i\\[0.1em]
 \qquad \tau_i, e_i, L_i = \sem{\sigma_i}_{e'},\;\; e'=\ident{getField}(\nu?, \nu_i, e)\\[0.6em]
\end{array}
\end{equation*}
%
% Collections
\begin{equation*}
\begin{array}{l}
 \sem{\,[\sigma]\,}_e = \ident{list}\langl\tau\rangl, \ident{List.map}~(\lambda x\rightarrow e')~(\ident{getChildren}(e)), L\\[0.4em]
 \quad \textnormal{where}~~\tau, e', L = \sem{\hat{\sigma}}_x
\end{array}
\end{equation*}

@@

Given a JSON, XML or CSV document, the F\# Data implementation constructs a structural value $s$
from the sample (this is straightforward, but Section~\ref{sec:impl-parsing} discusses some interesting 
aspects). The following defines a mapping $\semalt{-}$ which turns a sample value $s$ into a structural
type $\sigma$:
%
\begin{equation*}
\begin{array}{rclcrcl}
 \semalt{0} &\narrow{=}& \ident{bit} &\qquad& \semalt{1} &\narrow{=}& \ident{bit} \\
 \semalt{i} &\narrow{=}& \ident{int} && \semalt{d} &\narrow{=}& \ident{decimal} \\
 \semalt{f} &\narrow{=}& \ident{float} && \semalt{t} &\narrow{=}& \ident{string} \\
 \semalt{b} &\narrow{=}& \ident{bool} && \semalt{\kvd{null}}  &\narrow{=}& \kvd{null} \\[0em]
\end{array}
\end{equation*}
\noindent
\vspace{-0.2em}
\begin{equation*}
\begin{array}{l}
 \semalt{[s_1; \ldots; s_n]} = [\semalt{s_1, \ldots, s_n}]
 \\[0.5em]
 \semalt{\nu?~\{ \nu_1 \mapsto s_1, \ldots, \nu_n \mapsto s_n \}} =\\[0.1em]
 \qquad\nu?~\{ \nu_1:\semalt{s_1}, \ldots, \nu_n :\semalt{s_n} \}
 \\[0.5em]
 \semalt{s_1, \ldots, s_n} = \sigma_n \quad\textnormal{where}\\[0.1em]
 \qquad\sigma_0 = \top,~ \forall i\in \{ 1.. n \}.~ \sigma_{i-1} \triangledown \semalt{s_i} \vdash \sigma_i
\end{array}
\end{equation*}
%
Primitive values are mapped to their corresponding types, with the exception of $0$ and $1$, which
are inferred as \ident{bit}. For records, we return a type with field types inferred based on 
individual values.
